---
title: "Processing transcriptome annotations and transcript-level quantification using Salmon"
subtitle: ""
date: "`r Sys.Date()`"
output: 
    html_document:
        theme: readable
        toc: true
        toc_depth: 1
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

rm(list = ls())
```

# Background

Here, we investigate the impact of different choices of transcriptome
annotations on isoform-level quantifications from RNA-seq. We first
quantify RNA-seq data from [120 fetal human brain
samples](https://genomebiology.biomedcentral.com/articles/10.1186/s13059-018-1567-1),
using [Salmon](https://salmon.readthedocs.io/en/latest/index.html) with
various transcriptome annotations, such as different versions of GENCODE
and in-house annotations generated from single-cell Iso-seq (long-read)
data.

# Download GENCODE annotations

We download [the latest GENCODE
annotation](https://www.gencodegenes.org/human/) for the human genome
(GRCh37), which is Release 38. We download the comprehensive gene
annotation file in [GTF
format](https://uswest.ensembl.org/info/website/upload/gff.html). We
also download Relsease 33, which has been used by Cindy, and we use the
annotation files for GRCh37, not GRCh38, because GRCh37 has been
previously used by our lab members, and hence they are more suitable for
direct comparison, although it comes with [its own
caveats](http://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_human/_README_GRCh37_mapping.txt).
(***TODO's***: The effect of using [GRCh37 instead of
GRCh38](http://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_human/_README.TXT)
could be further studied. We could even quantify RNA-seq based on older
versions of GENCODE, such as v19 used by PsychENCODE or v26 used by
GTEx. RefSeq transcriptome annotation, which is largely a subset of
GENCODE annotations could be also investigated.)

```{r, eval=FALSE}
# Download GENCODE .gtf files
system("curl ftp.ebi.ac.uk/pub/databases/gencode/Gencode_human/release_38/GRCh37_mapping/gencode.v38lift37.annotation.gtf.gz --output GENCODE/hg37/gencode.v38lift37.annotation.gtf.gz")

system("curl ftp.ebi.ac.uk/pub/databases/gencode/Gencode_human/release_33/GRCh37_mapping/gencode.v33lift37.annotation.gtf.gz --output GENCODE/hg37/gencode.v33lift37.annotation.gtf.gz")
```

# Munge GENCODE annotations

In addition to using raw GENCODE GTF files to quantify RNA-seq data, we
created additional GENCODE files by filtering transcripts based on
`transcript support level`, where the score of 1 corresponds to best
line of evidence and the score of 5 less convincing evidence. This
information is contained in the last column of GTF file. Only
transcripts with `transcript support level` less than three were
included. (***TODO's***: We could potentially apply additional filters,
such as including only manually annotated HAVANA transcripts or
transcripts with [tag
"basic"](https://www.gencodegenes.org/pages/tags.html), but such filters
can lead to substantially smaller number of transcripts. For example,
only selecting transcripts with
[CAGE_supported_TSS](https://www.gencodegenes.org/pages/tags.html) would
leave only about 2,000 transcripts, which is far too small of a number.)

```{r, eval=FALSE}
library(tidyverse)
annot <- read_delim("GENCODE/hg37/gencode.v38lift37.annotation.gtf.gz", 
                    delim = "\t", skip = 5, col_names = FALSE)
dim(annot)
colnames(annot) <- c("chromosome", "source", "feature", "start", "end", 
                     "score", "strand", "phase", "info")

# Extract transcript ID
annot$transcript_id <- str_extract(annot$info, 
                                   "transcript_id (.*?);") %>% 
  str_remove("transcript_id \"") %>% 
  str_remove("\";")

annot_transcript <- annot %>% 
  filter(feature == "transcript")
# 238,613 transcripts for GENCODE v38
# 229,692 transcripts for GENCODE v33

# Extract transcript support level info
annot_transcript$tsl <- str_extract(annot_transcript$info, 
                                    "transcript_support_level (.*?);") %>% 
  str_remove("transcript_support_level ") %>%
  str_remove("\"") %>%
  str_remove("\"") %>% 
  str_remove(";")

annot_transcript %>% 
  count(tsl)

# Extract gene ID
annot_transcript$gene_id <- str_extract(annot_transcript$info, 
                                        "gene_id (.*?);") %>% 
  str_remove("gene_id \"") %>% 
  str_remove("\";")

# Extract gene name
annot_transcript$gene_name <- str_extract(annot_transcript$info, 
                                        "gene_name (.*?);") %>% 
  str_remove("gene_name \"") %>% 
  str_remove("\";")

# Extract transcript name
annot_transcript$transcript_name <- str_extract(annot_transcript$info, 
                                        "transcript_name (.*?);") %>% 
  str_remove("transcript_name \"") %>% 
  str_remove("\";")

annot_transcript$level <- str_extract(annot_transcript$info, 
                                        "level (.*?);") %>% 
  str_remove("level") %>%
  str_remove(";")

# Extract transcript type
annot_transcript$transcript_type <- str_extract(annot_transcript$info, 
                                        "transcript_type (.*?);") %>% 
  str_remove("transcript_type \"") %>%
  str_remove("\";")

annot_transcript$tag <- str_extract(annot_transcript$info, 
                                        "tag (.*?);") %>% 
  str_remove("tag \"") %>%
  str_remove("\";")

annot_transcript_filter <- annot_transcript %>% 
  filter(tsl %in% c(1, 2, 3))
# 113,308 transcripts for GENCODE v38
# 113,878 transcripts for GENCODE v33

xtabs(~ level + tsl, annot_transcript)
cor.test(as.numeric(annot_transcript$tsl), as.numeric(annot_transcript$level))
xtabs(~ transcript_type + tsl, annot_transcript)
xtabs(~ level + tsl, annot_transcript_filter)
xtabs(~ tag + tsl, annot_transcript_filter)

annot %>% 
  filter(feature %in% c("transcript", "exon")) %>% 
  semi_join(annot_transcript_filter, by = "transcript_id") %>% 
  select(-10) %>% 
  write_tsv(file = "gencode.v38lift37.annotation.filter.gtf", 
            col_names = FALSE, quote = "none", escape = "none")
# write_tsv can also write .gz file,
# but here we save it as .gtf to use it as an input to `gffread`
```

# Munge Ashok's single-cell long-read annotations

Different transcriptome annotations generated from Ashok's long-read are
also used for transcript-level quantification.
`ashok-isoform.annotation.pseudobulk.gtf.gz` contains all transcripts
supported by three or more reads (\~170k isoforms), while
`ashok-isoform.annotation.7189.allsinglecells.gtf.gz` contains isoforms
with UMI \> 85 (\~140k isoforms).
`ashok-isoform.annotation.7189.allsinglecells.gtf.gz` is a smaller
subset of `ashok-isoform.annotation.pseudobulk.gtf.gz`. Finally,
`ashok-gencode.v33lift37.pseudobulk.gtf.gz` combines GENCODE v33 with
`ashok-isoform.annotation.pseudobulk.gtf.gz`. Each of these three
annotations are used to run Salmon.

```{r, eval=FALSE}
annot <- read_delim("GENCODE/hg37/ashok-gencode.v33lift37.pseudobulk.gtf.gz", 
                    delim = "\t", skip = 5, col_names = FALSE)
dim(annot)
colnames(annot) <- c("chromosome", "source", "feature", "start", "end", 
                     "score", "strand", "phase", "info")

annot$transcript_id <- str_extract(annot$info, 
                                   "transcript_id (.*?);") %>% 
  str_remove("transcript_id \"") %>% 
  str_remove("\";")

annot %>% 
  count(feature)
# 142,573 transcripts for isoform.annotation.7189.allsinglecells
# 173,782 transcripts for isoform.annotation.pseudobulk
# 355,861 transcripts for gencode.v33lift37.pseudobulk

annot %>% 
  filter(feature %in% c("transcript", "exon")) %>% 
  select(-10) %>% 
  write_tsv(file = "ashok-gencode.v33lift37.pseudobulk.transcripts.gtf", 
            col_names = FALSE, quote = "none", escape = "none")
```

# Create transcriptome fasta files for Salmon indices

We use
[`gffread`](http://ccb.jhu.edu/software/stringtie/gff.shtml#gffread_ex)
to extract transcript sequences from the GRCh37 fasta file. Such
transcript fasta files are subsequently used to create [Salmon
indices](https://salmon.readthedocs.io/en/latest/salmon.html#preparing-transcriptome-indices-mapping-based-mode),
which are used for pseudoalignment.

```{bash, eval=FALSE}
gffread-0.12.6.Linux_x86_64/gffread GENCODE/hg37/gencode.v33lift37.annotation.filter.gtf -g GENCODE/hg37/GRCh37.primary_assembly.genome.fa -w GENCODE/hg37/gencode.v33lift37.filter.transcripts.fa

gzip GENCODE/hg37/gencode.v33lift37.filter.transcripts.fa

cat GENCODE/hg37/gencode.v33lift37.filter.transcripts.fa.gz GENCODE/hg37/GRCh37.primary_assembly.genome.fa.gz > salmon_index/gencode.v33lift37.filter.transcripts.decoy.fa.gz

salmon-1.5.1_linux_x86_64/bin/salmon index -t salmon_index/gencode.v33lift37.filter.transcripts.decoy.fa.gz -d salmon_index/decoys.txt -p 4 -i salmon_index/gencode.v33lift37.filter.salmon_index --gencode
```

# Quantify using Salmon

We usedthe following bash script to run Salmon and hence quantify
transcript-level counts. (***TODO's***: Salmon quantifications should be
compared against STAR + RSEM quantifcations, which use a similar
expectation-maximization (EM)-based optimization.)

```{bash, eval=FALSE}
#!/bin/bash
#$ -l h_data=4G,h_rt=12:00:00,highp
#$ -cwd
#$ -pe shared 12
#$ -j y
#$ -o ./log/obrien.out
#$ -m a
#$ -t 1-120
cd /u/project/gandalm/shared/bigsummer
ANNOT=gencode.v33lift37.filter
SALMON_INDEX=./salmon_index/${ANNOT}.salmon_index
INPUT_FILE=/u/project/gandalm/cindywen/isoform_twas/star/sample_list_obrien.txt
INLINE=`head -n ${SGE_TASK_ID} ${INPUT_FILE} | tail -n 1`
IFS=$'\t' 
params=(${INLINE})
SAMPLE_ID=${params[0]}
IN_FASTQS_R1=${params[1]}
IN_FASTQS_R2=${params[2]}
SAMPLE_PATH=./obrien/${ANNOT}/${SAMPLE_ID}
mkdir -p ${SAMPLE_PATH}
./salmon-1.5.1_linux_x86_64/bin/salmon quant -i ${SALMON_INDEX} -l A -1 ${IN_FASTQS_R1} -2 ${IN_FASTQS_R2} --validateMappings --useEM --seqBias --gcBias -p 12 -o ${SAMPLE_PATH}
```

# Load transcript-level quantification

We read in quantification result for one sample as a sanity check.

```{r, eval=FALSE}
sample <- read_delim("obrien/gencode.v33lift37.filter/1117/quant.sf", delim = "\t")
sample %>% 
  mutate(normalized = NumReads / EffectiveLength) %>% 
  mutate(estimateTPM = normalized / sum(normalized) * 1e6) %>% 
  head()
# some degree of discordance between estimated TPM and TPM of Salmon output, 
# not sure why
```
